Forage site distributions quantify collective foraging in honey bee colonies

### 2021-11-16 analysis/manuscript/paper.Rmd
updated manuscript and sent to Vincent and Elli

# Main

Collective decision making provides animals groups with the ability to pool individual information and make informed decisions. Despite the enhanced ability to process information, not every circumstance requires a collective response. In many cases, animal groups do better when individuals prioritise their own personal information over social information provided by conspecifics [@King2007; @Rieucau2011; @Gruter2013]. Identifying the conditions which favour different information use strategies remains a key challenge, both for determining what conditions likely lead to the evolution of collective decision making and for understanding how animal groups might interact in an increasingly anthropomorphic world. Nevertheless, assessing the importance of different information use strategies can be extremely difficult, requiring the actions and decision making of a collective to be compared with that of individual actors. This is fundamentally difficult as making such comparisons demands observations of individual and collective behaviours. A side from limiting the group size that can be studied, in many cases we can observe one, or the other, but not both.

In honey bees, foraging is collectively coordinated via the famous "waggle dance", which conveys the location of a profitable resource as a vector of distance (duration of the waggle run) and direction (angle of the dance) [@Frisch1993]. Resource “quality” is not encoded in the form of the dance itself, but in the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Hasenjager2021], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995]. This creates a feedback mechanism through which the hive processes the information collected and proportionally selects the most profitable resources [@Seeley1994A].

Despite the efficiency benefits the waggle dance reportedly provides, its contribution to colony foraging success remains unclear. In many situations, colonies that are prevented from communicating via dance behaviour achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002; @Dornhaus2004; @Price2019]. In some ecological circumstances, individuals that forgo waiting for dance information in favour of individual search, termed “scouting”, may find food more quickly, and evidence suggests that bees adopt this strategy if the benefits of dance-following drop [@Gruter2008]. Overall, these studies suggest that honey bees use the waggle dance flexibly in response to environmental conditions [@Guter2011; @Gruter2014]. However, it is difficult to empirically probe this relationship further without new methods which can tease apart the extent to which foraging occurs with or without use of the dance.

Here, we describe a model to quantify waggle dance use in honey bee foraging and fit it to a large empirical waggle dance dataset. After estimating the extent to which colony foraging occurs individually or collectively, we investigate the influence different land-use types have on waggle dance use. Our methods provide novel evidence for collective honey bee foraging and a framework to evaluate how environmental conditions influence waggle dance use in honey bees.

Whether an individual forager follows a waggle dance from the dance floor or finds a resource for themselves, all foragers report resource information on the dance-floor. Consequently, the difference in foraging behaviours between individuals using the dance and individuals relying on personal information should manifest itself in the pattern of dances reported on the dance-floor. To evaluate this idea, we simulated honey bees foraging in a landscape using both personal and waggle dance information (Fig. 1). As which individuals foraged under what strategy is known in the simulation, we can compare the distributions of foraging distances reported on the dance floor. Figs 1c and 1d show that the shapes of the resource distance distributions under the two simulated scenarios are very different. The distribution of the first simulation is akin to that of an exponential distribution (Fig. 1 C), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported by recruits (Fig. 1 D) is appears to resemble a Rayleigh distribution which is the nearest neighbour distribution of a two-dimensional environment (see methods).

@@

As well as providing direct evidence that honey bees forage collectively according to the widely accepted theory (refs), our results demonstrate that waggle dance use depends on environmental context. Although this has been suspected for some time (refs), empirical evidence that colonies foraging on the landscape might regulate the use of the waggle dance have relied on the method of dance disorientation (refs for dance disorientation study), whereby a colony is disoriented to remove the spatial component of the waggle dance. Foraging success, measured as quantities of nectar or pollen retrieved, is then compared to an undisturbed colony in the same environment to determine what dance use strategy is the most effective. Whilst this method allows for an evaluation of whether or not colonies forage more effectively with or without the dance in a given environment, whether a colony actively regulates waggle dance use in response to the environment is not directly shown. In showing that foraging patterns in three of our sites are best explained by individual foraging, our results demonstrate honey bee colonies tune waggle dance use in relation to environmental factors.

As our model predicts the proportion of dances coming from an individual or collective foraging strategy, its fit to waggle dance data provides a tool to estimate a colonies use of the waggle dance.

@@

In the Agri-rural environment a single PC explained ~60% of the variation in the proportion of scouts (beta regression: $R^2$ = 0.6, $\phi$ = 4.9, p < 0.05, Fig. 4). This PC correlates positively with arable land and negatively with non-agricultural unmanaged green space, built up areas and water. Arable land is typically nutritionally poor for honey bees (ref). On the contrary, non-agricultural unmanaged green space is typically nutrient rich, as are built up areas in agri-rural environments, which often have gardens with ample flowers available for pollinators (ref). Water also correlated negatively with the estimated proportion of scouts, possibly due to river and pond banks providing good floral resources for pollinators (ref). Combined, these results suggests when resources are sparse and difficult to find, as with arable land-use, individuals value personal information more than when easier to find, as with the non-agricultural unmanaged green space, built up areas and water.

@@

A similar effect is also observed in the urban environment, where a single PC explained ~61% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.61, $\phi$ = 7.4, p < 0.05, Fig. 4). This PC correlates positively with continuos central land, dense residential land and water, whilst correlating negatively with sparse residential and amenity grassland. This relationship appears to be largely driven by a single site showing a model of only individual foraging as the best fit. As with the agri-rural land-uses, these results indicate when resources are sparse, as could be expected in continuos central and dense residential land, where there is limited space for floral resources, a colony emphasises the use of personal rather than social information. However, when resources are more abundant, as could be expected in sparse residential areas and amenity grasslands, a colony appears to increase the use of information contained in the waggle dance. Unlike in the agri-rural environment, water correlates positively with a higher use of the waggle dance. This likely indicates that the banks of urban water ways are less nutritionally rich than in agri-rural landscapes.

@@

In both the agri-rural and Urban environments, land-use types which associate with resource scarcity correlate positively with a higher proportion of scouts, whilst those which correlate with resource abundance correlate with a lower proportion of scouts, indicating a greater use of the waggle dance. These results appear to follow what is predicted under theoretical models: in landscapes where resources are likely to more difficult to find, a colony tends to have a higher number of scouts than in landscapes where resources should be easier to locate [@Dornhaus2006; @Beekman2008]. However, it is important to note that other combinations which didn't show as having a significant association with the first principle component could have just as well fit into this argument, arguably better than others. In both the agri-rural and urban environments water seems to correlate in contrasting ways with the proportion of scouts. This could be due to differences in the floral richness of urban and agri-rural water ways (reff), however, if land-use type is to be taken as a proxy for floral availability, other variables such as parks, allotments and cemeteries in the urban environment and fruit or Oil Seed Rape in the agri-rural environment might be expected to be included in the first PC over water. An alternative explanation could be that water acts as barriers between land-use types and so gets bundled into the association (more thought on this needed i think).

To what extent social information is used when making foraging decisions, as well as the value of personal information, has been a matter of considerable debate (refs). Much of that debate is informed by simulation studies (ref), and empirical work in which a relatively small number of individuals are observed (ref). Here, we demonstrate that the analysis of waggle dance data can make an important contribution to this debate and provide a methodology to further evaluate how honey bees use this unique behaviour. With the advances in the decoding of the waggle dance through automated methods [@Wario2017], we face the prospect of waggle dance data becoming “big data”. Our methodology provides a means of analysing such large data sets and gleaning useful information on the importance of such data to inform the debate about the importance of social information, as well as providing useful colony metrics of foraging activity.

>>>>

# Main (Word count excluding figures and legend = 2357)

Collective decision making provides animals groups with the ability to pool individual experiences and make more informed decisions. Despite an enhanced capacity to process information, not every circumstance requires a collective response. In many cases, animal groups have been shown to do better when individuals prioritise information they obtain through their own experiences to that provided by conspecifics [@King2007; @Rieucau2011; @Gruter2013]. Identifying the conditions which favour different information use strategies is therefore important, both for determining what conditions might have lead to the evolution of sociality and for understanding how animal groups might interact in an increasingly anthropomorphic world. However, assessing the importance of different information use strategies is not trivial and requires comparing the actions and behaviours of individuals with that of a collective. Often, information from different sources gets pooled together and so determining the importance of social Vs individual information use requires tools which can tease apart individual and collective behaviours.

Honeybees provide an ideal study system in which to evaluate social information use thanks to the famous "waggle dance", which conveys the location of a profitable resource as a vector of distance (duration of the waggle run) and direction (angle of the dance) [@Frisch1993]. Resource “quality” is not encoded in the form of the dance itself, but in the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Hasenjager2021], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995]. This creates a feedback mechanism through which the hive processes the information collected and proportionally selects the most profitable resources [@Seeley1994A].

Despite the efficiency benefits the waggle dance reportedly provides, its contribution to colony foraging success remains unclear. In many situations, colonies that are prevented from communicating via dance behaviour achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002; @Dornhaus2004; @Price2019]. In some ecological circumstances, individuals that forgo waiting for dance information in favour of individual search, termed “scouting”, may find food more quickly, and evidence suggests that bees adopt this strategy if the benefits of dance-following drop [@Gruter2008]. Whilst individuals can be observed under experimental conditions to determine how they use different sources of information, translating this up to a colony level is logistically impossible due to honey bees small size, extensive foraging ranges and shear number of foragers. New methods which can quantify waggle dance use at a colony level are therefore needed if we are to assess the conditions where the dance provides the most benefit and derive the conditions user which this unique behaviour evolved.

Whether an individual forager follows a waggle dance from the dance floor or finds a resource for themselves, all foragers report resource information on the dance-floor [@Seeley1995]. Consequently, the difference in foraging behaviours between individuals following the dance and individuals relying on personal information should manifest itself in the pattern of dances reported on the dance-floor. To evaluate this idea, we simulated honey bees foraging in a landscape using both personal and waggle dance information (Fig. 1). As which individuals forage under what strategy is known in the simulation, we can compare the distributions of foraging distances reported on the dance floor. Figs 1c and 1d show that the shapes of the resource distance distributions under the two simulated scenarios are very different. The distribution of the first simulation is akin to that of an exponential distribution (Fig. 1 C), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported by recruits (Fig. 1 D) is appears to resemble a Rayleigh distribution which is the nearest neighbour distribution of a two-dimensional environment (see methods).

@@

As well as providing direct evidence that honey bees forage collectively according to a widely accepted theory [@Seeley1994; @Sherman2002], in showing that foraging patterns in three of our sites are best explained by individual foraging, our results demonstrate honey bee colonies tune waggle dance use in relation to environmental factors. These results support previous efforts to quantify collective foraging [@Sherman2002; @Dornhaus2006; @Gruter2013; @Price2019]. Furthermore, as our model predicts the proportion of dances coming from an individual or social foraging strategy, its fit to waggle dance data provides a tool to estimate a colonies use of the waggle dance.

@@

In the Agri-rural environment a single PC explained ~60% of the variation in the proportion of scouts (beta regression: $R^2$ = 0.6, $\phi$ = 4.9, p < 0.05, Fig. 4). This PC correlates positively with arable land and negatively with non-agricultural unmanaged green space, built up areas and water. Arable land is typically nutritionally poor for honey bees (ref). On the contrary, non-agricultural unmanaged green space is typically nutrient rich, as are built up areas in agri-rural environments, which often have gardens with ample flowers available for pollinators (ref). Water also correlated negatively with the estimated proportion of scouts, possibly due to river and pond banks providing good floral resources for pollinators (ref). Combined, these results suggest when resources are sparse and difficult to find, as with arable land-use, individuals value personal information more than when easier to find, as with the non-agricultural unmanaged green space, built up areas and water.

@@

A similar effect is also observed in the urban environment, where a single PC explained ~61% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.61, $\phi$ = 7.4, p < 0.05, Fig. 4). This PC correlates positively with continuous central land, dense residential land and water, whilst correlating negatively with sparse residential and amenity grassland. This relationship appears to be largely driven by a single site showing a model of only individual foraging as the best fit. As with the agri-rural land-uses, these results indicate when resources are sparse, as could be expected in continuous central and dense residential land, where there is limited space for floral resources, a colony emphasises the use of personal rather than social information. However, when resources are more abundant, as could be expected in sparse residential areas and amenity grasslands, a colony appears to increase the use of information contained in the waggle dance. Unlike in the agri-rural environment, water correlates positively with a higher use of the waggle dance. This likely indicates that the banks of urban water ways are less nutritionally rich than in agri-rural landscapes.

@@

In both the agri-rural and Urban environments, land-use types which associate with resource scarcity correlate positively with a higher proportion of scouts, whilst those which correlate with resource abundance correlate with a lower proportion of scouts, indicating a greater use of the waggle dance. These results appear to follow what is predicted under theoretical models: in landscapes where resources are likely to more difficult to find, a colony tends to have a higher number of scouts than in landscapes where resources should be easier to locate [@Dornhaus2006; @Price2019]. However, it is important to note that other combinations which didn't show as having a significant association with the first principle component could have just as well fit into this argument, arguably better than others. In both the agri-rural and urban environments water seems to correlate in contrasting ways with the proportion of scouts. This could be due to differences in the floral richness of urban and agri-rural water ways (reff), however, if land-use type is to be taken as a proxy for floral availability, other variables such as parks, allotments and cemeteries in the urban environment and fruit or Oil Seed Rape in the agri-rural environment might be expected to be included in the first PC over water. An alternative explanation could be that water acts as barriers between land-use types and so gets bundled into the association (more thought on this needed I think).

Here, we demonstrate that the analysis of waggle dance data can make an important contribution to our understanding of social information use and provide a methodology to further evaluate how honey bees use this unique behaviour. Recently, technological advances have emerged which enable colony metrics to be collected faster, more accurately and over greater time spans than could be gathered by hand .

Recent advances in computational methods have allowed individuals to be tracked within colonies in ways which would be almost impossible to do by hand, allowing theories of individual behaviour to be evaluated in more depth than could have been done previously [@Wild2021]. We are not yet in a position to be able to track behaviours outside the colony in the same detail, nevertheless, by piecing together the behavioural response of individuals into a mathematical description of colony foraging, the model outlined here extends our ability to predict individual behaviour beyond the colony walls and in to the environment. With the advances in the decoding of the waggle dance through automated methods [@Wario2017], we face the prospect of waggle dance data becoming “big data”. Our methodology provides a means of analysing such large data sets to inform the debate about the importance of social information, as well as providing useful colony metrics of foraging activity.
